Leadership across cultures and contexts: an exploratory analysis

### 2020-05-01 leadership_across_cultures_contexts.Rmd
Minor changes in intro section.
Added to discussion -- status results.

Research on leadership within anthropology extends to the birth of the discipline. Early on, anthropologists (and some sociologists) realized that leadership and followership were critical to understanding human psychology, social organization, and culture [e.g., @mumford_origins_1906; @firth1927economic; @morgan_ancient_1877; @myres_influence_1917]. Anthropologists have documented some sort of leadership among every ethnographically observed society [@brown_human_1991; @lewis_leaders_1974], and across many social contexts, such as families and kin groups [@dussart2000politics], ritual [@singh_cultural_2017], work groups [@macfarlan_competitive_2012], conflicts between groups [@glowacki_evolutionary_2017], and in nonindustrial political groups [@cohen_comparative_1967]. For review, see @garfield_review_2019.

Leaders substantially shape group structures and outcomes across cultures. The structure of groups and their goals, in turn, as well as their subsistence strategies, can favor particular leader qualities. Despite decades of research and theorizing about leadership in numerous social science disciplines, however, to our knowledge there is no comprehensive or systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of nonindustrial societies. Moreover, outside of anthropology, most research and theorizing on leadership is based on data from Western or Westernized, postindustrial societies [@von_rueden_leadership_2015].

@@

There is now well established consensus within the behavioral sciences on the importance of data from the full range of human cultural diversity. Systemic biases in much of social science are well known, and might be due, in part, to the lack of data from populations other than American undergaduates [@henrich_weirdest_2010, *The weirdest people in the world?*].

@@

There is as much diversity within non-Western cultures as there is within Western ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-realiance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

Extending the study of leadership to include evolutionary theory and data from diverse, small-scale and traditional societies (i.e., non-WEIRD) can facilitate more generalizable theories and offer applications to shape organizational structure across populations (i.e., WEIRD settings) in ways which are more consistent and less mismatched with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. Such approaches can also incorporate the anthropological insight that diversity of cultural, social, and ecological contexts are strong drivers of human behavior and group dynamics [@wiessner_risk_1982; @henrich_big_2015].

@@

<!-- EVOLUTIONARY THEORY APPROACH -->
Much scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover the interrelationships of these dimensions of leadership. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @wiessner_collective_2019; @redhead_dynamics_2018; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020].

<!-- GROUP CONTEXT and cost PROBLEMS -->
Much of leadership studies, including evolutionary approaches, have focused on community or political leaders, often overlooking subtilties in the nature of leadership and followership across social domains and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. Lastly, while the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated. Outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is associated with population-level variation and in different social contexts and identifying the associated benefits *and* costs both leaders and followers incur.
<!-- CAN PULL SOME COST CITATIONS FROM NHB PAPER -->

<!-- THIS PARAGRAPH CAN BE INCORPORATED ABOVE ????? -->
Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. Leadership roles are also often associated with social sanctioned rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @von_rueden_leadership_2015].

@@

<!-- MALE BIAS -->
Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse non-human social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 demonstrated in a systematic cross-cultural sample that political leaders of both sexes were present in only seven of 93 societies and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group and women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. Empirical ethnographic studies have suggested sex-differences in political influence and leadership across human societies including among more egalitarian cultural contexts are more likely driven by ecological and social structural factors, rather than sex (or gender identity) per se, including social capital and access to same-sex allies [@yanca_female_2004] and/or sex differences in the sexual division of labor [@von_rueden_sex_2018]. At least two studies relying on ethnographic fieldwork comparing leaders and non-leaders in relativley small-scale societies revealed sex is not a strong predictor of community influence or elected leadership positions, after accounting for a broad range of demographic, phenotypic, cognitive, social, and other traits derived from the evolutionary theoretical literature [e.g., @garfield_investigating_2019; @von_rueden_sex_2018].

@@

In sum, we sought to empirically and systematically accomplish the following four goals: 1) identify variation in the *qualities of leaders* and the *functions they serve*. 2) Identify *features* of the qualities and functions of leaders (e.g., do particular qualities or functions tend to covary or cluster together?) in reference to theoretical models or novel interpretations. Such features might lend support to existing theories; for instance they might correspond to qualities associated with "prestige" or "dominance" [@henrich_evolution_2001] or functions associated with conflict or cooperation [@gavrilets_collective_2015; @glowacki_evolutionary_2017; @hooper_theory_2010]. 3) Identify variation in the *costs and benefits* of leadership for both leaders and followers. And 4) determine the extent to which the qualities and functions of leaders, and their costs and benefits, varied across social contexts within populations, varied across popululations with different subsistence strategies, and varied by continental region.

@@

In constrast to the qualities and functions variables, coding operationalizations of the costs and benefits of leadership were developed *a priori* based on our reading of the evolutionary theoretical literature from anthropology and biology. For both followers and leaders, we coded if there was evidence of costs paid or benefits received in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories. See Table \@ref(tab:costbenefitvars). There are multiple instances of conceptual overlap between many of our researcher-coded variables, in particular between various qualities and related costs or benefits. The operationalization of costs and benefits generally capture a wider range of evidence than related qualities and focus more on evidence of transaction rather than description. We return to this point in the limitations section.

@@

We analyzed data from our researcher-coded operationalized variables at the text-record level and at the culture-level. Our first goal was to assess the proportion of text records and cultures that provided evidence for each of our coded variables to discover the qualities, functions, benefits and costs of leadership that are most and least frequently noted in the ethnographic record. We computed the proportion of text records supporting each variable, and their 95% confidence intervals, using logistic generalized linear mixed effects models with random intercepts for document authors nested within cultures (which partially accounts for potential variation due to differences in ethnographers such as gender, nationality, or theoretical framework), controlling for one potential source of bias as a fixed effect: the date of publication. For each varible, we then computed the proportion of cultures with at least one text record with evidence for that variable, with confidence intervals computed using a cluster bootstrap.

@@

Susistence strategy was confounded with continental region. African text records were predominantly from horticultural societies, for instance, whereas North American text records were prodominantly from hunter-gatherer societies (see Figure \@ref(fig:regionsubsistence)).

Susistence strategy and group structure were also confounded. Hunter-gatherer text records often refered to residential leadership (`r hg_residential_pct`% of the total sample of hunter-gatherer text records) and kin leadership (`r hg_kin_pct`%), but rarely refered to supracommunity leadership (`r hg_supra_pct`%). Agricultural text records, on the other handj, rarely referring to residental leadership (`r agri_residential_pct`%) but often referring to supracommunity leadership (`r agri_supra_pct`%). Across all subsistence types, the most frequent group type was political groups at the community level, with the exception of agriculturalists in which supracommunity political groups are the most frequent group type. Leadership beyond the community level is relatively rare outside of agricultural populations and is especially rare among hunter-gatherers. See Figure \@ref(fig:grouptype).

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded variables (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support using logistic mixed effects models that controlled for one possible source of bias -- pubication date -- as a fixed effect, and included authors nested within cultures as random intercepts. We then computed the proportion of cultures with at least one text record providing support for that variable (calculating 95% CI using a cluster bootstrap). See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CI's, see Supplementary Table S??.

@@

*High status* was the most frequent leader quality in the ethnographic record, with `r statusPercTxt`% of text records and `r statusPercCult`% cultures containing evidence that leaders were of high social status. To explore factors that distinguished these records from those with no evidence for status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see SI for details). We then fit an elasticnet logistic regression model (with lasso penalty) of high status (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provide evidence of high status as a quality of leaders. These included *`r rev(highstatus_plot$data$Word)[1:4]`*, and *`r rev(highstatus_plot$data$Word)[5]`*. See Figure \@ref(fig:highstatusplot).

We then investigated if the proportion of text records with evidence for high status differed among the different levels of social and political organization, subsistence strategies, continental regions, and leader sex (again using a logistic mixed effects model with these variables as fixed effects and authors nested within cultures as random intercepts). There were modest differences. Hunter-gatherers and pastorilist societies were more likely to provide evidence for high status, as was South America and East Asia. Female leaders were also more likely to be identified as high status. Somewhat counter-intuitively, the proportion of text records indicating high status was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus). For the table of coefficients, see Supplementary Table X.

@@

@@

Leadership and high social status are expected to be associated with particular suites of characteristics, which potentially vary by cultural or social context, personality, or via other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001]. Exploratory cluster analyses reported here suggest multiple leadership phenotypes.

@@

## Disentangling the relationship between leadership and high status

>>>>

Research on leadership within anthropology extends to the birth of the discipline. Early on, anthropologists (and some sociologists) realized that leadership and followership were critical to understanding human psychology, social organization, and culture [e.g., @mumford_origins_1906; @firth1927economic; @morgan_ancient_1877; @myres_influence_1917]. Anthropologists have documented the emergence of leaders among every ethnographically observed society [@brown_human_1991; @lewis_leaders_1974], and across diverse social contexts, such as families and kin groups [@dussart2000politics], ritual [@singh_cultural_2017], work groups [@macfarlan_competitive_2012], conflicts between groups [@glowacki_evolutionary_2017], and in nonindustrial political groups [@cohen_comparative_1967]. For review, see @garfield_review_2019.

Leaders substantially shape group structures and outcomes across cultures. The structure of groups and their goals, in turn, as well as their subsistence strategies, can favor particular leader qualities. Despite decades of research and theorizing on leadership in numerous social science disciplines, however, to our knowledge there is no comprehensive or systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of nonindustrial societies. Moreover, outside of anthropology, most research and theorizing on leadership is based on data from Western or Westernized, postindustrial societies [@von_rueden_leadership_2015].

@@

There is now well established consensus within the behavioral sciences on the importance of data from the full range of human cultural diversity. Systemic biases in much of social science are well known, and might be due, in part, to the lack of data from populations other than American undergraduates [@henrich_weirdest_2010, *The weirdest people in the world?*].

@@

There is potentially as much diversity within nonindustrial, non-WEIRD cultures as there is within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

Extending the study of leadership to include evolutionary theory and data from diverse, small-scale and nonindustrial societies (i.e., non-WEIRD) can facilitate more generalizable theories and offer applications to shape organizational structure across populations (i.e., WEIRD settings) in ways which are more consistent and less mismatched with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. Such approaches can also incorporate the anthropological insight that diversity of cultural, social, and ecological contexts are strong drivers of human behavior and group dynamics [@wiessner_risk_1982; @henrich_big_2015].

@@

<!-- EVOLUTIONARY THEORY APPROACH -->
Substantial scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover interrelationships of dimensions of leadership, including the traits of leaders, their roles, and relationships with culture, economics, and group structure. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @wiessner_collective_2019; @redhead_dynamics_2018; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020].

<!-- GROUP CONTEXT and cost PROBLEMS -->
Much of leadership studies, including evolutionary approaches, have focused on community or political leaders, often overlooking subtleties in the nature of leadership and followership across social contexts and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. Lastly, while the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated [for discussions on costs of leadership see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014]. Outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is associated with between-group population-level variation and within-group variation in social contexts, as well as identifying the associated benefits *and* costs both leaders and followers incur.
<!-- CAN PULL SOME COST CITATIONS FROM NHB PAPER -->

<!-- THIS PARAGRAPH CAN BE INCORPORATED ABOVE ????? -->
Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. Leadership roles are also often associated with socially sanctioned rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @von_rueden_leadership_2015].

@@

<!-- MALE BIAS -->
Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse non-human social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 demonstrated in a systematic cross-cultural sample that political leaders of both sexes were present in only seven of 93 societies and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group and women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. Empirical ethnographic studies have suggested sex-differences in political influence and leadership across human societies including among more egalitarian cultural contexts are more likely driven by ecological and social structural factors, rather than sex (or gender identity) per se, including social capital and access to same-sex allies [@yanca_female_2004] and/or sex differences in the sexual division of labor [@von_rueden_sex_2018]. At least two studies relying on ethnographic fieldwork comparing leaders and non-leaders in relatively small-scale societies revealed sex is not a strong predictor of community influence or elected leadership positions, after accounting for a broad range of demographic, phenotypic, cognitive, social, and other traits derived from the evolutionary theoretical literature [e.g., @garfield_investigating_2019; @von_rueden_sex_2018].

@@

In sum, we sought to empirically and systematically accomplish the following four goals: 1) identify variation in the *qualities of leaders* and the *functions they serve*. 2) Identify *features* of the qualities and functions of leaders (e.g., do particular qualities or functions tend to covary or cluster together?) in reference to theoretical models or novel interpretations. Such features might lend support to existing theories; for instance they might correspond to qualities associated with "prestige" or "dominance" [@henrich_evolution_2001] or functions associated with conflict or cooperation [@gavrilets_collective_2015; @glowacki_evolutionary_2017; @hooper_theory_2010]. 3) Identify variation in the *costs and benefits* of leadership for both leaders and followers. And 4) determine the extent to which the qualities and functions of leaders, and their costs and benefits, varied across social contexts within populations, varied across populations with different subsistence strategies, and varied by continental region.

@@

In contrast to the qualities and functions variables, coding operationalizations of the costs and benefits of leadership were developed *a priori* based on our reading of the evolutionary theoretical literature from anthropology and biology. For both followers and leaders, we coded if there was evidence of costs paid or benefits received in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories. See Table \@ref(tab:costbenefitvars). There are multiple instances of conceptual overlap between many of our researcher-coded variables, in particular between various qualities and related costs or benefits. The operationalization of costs and benefits generally capture a wider range of evidence than related qualities and focus more on evidence of transaction rather than description. We return to this point in the limitations section.

@@

We analyzed data from our researcher-coded operationalized variables at the text-record level and at the culture-level. Our first goal was to assess the proportion of text records and cultures that provided evidence for each of our coded variables to discover the qualities, functions, benefits and costs of leadership that are most and least frequently noted in the ethnographic record. We computed the proportion of text records supporting each variable, and their 95% confidence intervals, using logistic generalized linear mixed effects models with random intercepts for document authors nested within cultures (which partially accounts for potential variation due to differences in ethnographers such as gender, nationality, or theoretical framework), controlling for one potential source of bias as a fixed effect: the date of publication. For each variable, we then computed the proportion of cultures with at least one text record with evidence for that variable, with confidence intervals computed using a cluster bootstrap.

@@

Subsistence strategy was confounded with continental region. African text records were predominantly from horticultural societies, for instance, whereas North American text records were predominantly from hunter-gatherer societies (see Figure \@ref(fig:regionsubsistence)).

Subsistence strategy and group structure were also confounded. Hunter-gatherer text records often referred to residential leadership (`r hg_residential_pct`% of the total sample of hunter-gatherer text records) and kin leadership (`r hg_kin_pct`%), but rarely referred to supracommunity leadership (`r hg_supra_pct`%). Agricultural text records, on the other hand, rarely referring to residential leadership (`r agri_residential_pct`%) but often referring to supracommunity leadership (`r agri_supra_pct`%). Across all subsistence types, the most frequent group type was political groups at the community level, with the exception of agriculturalists in which supracommunity political groups are the most frequent group type. Leadership beyond the community level is relatively rare outside of agricultural populations and is especially rare among hunter-gatherers. See Figure \@ref(fig:grouptype).

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded variables (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support using logistic mixed effects models that controlled for one possible source of bias -- publication date -- as a fixed effect, and included authors nested within cultures as random intercepts. We then computed the proportion of cultures with at least one text record providing support for that variable (calculating 95% CI using a cluster bootstrap). See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CIs, see Supplementary Table S??.

@@

*High status* was the most frequent leader quality in the ethnographic record, with `r statusPercTxt`% of text records and `r statusPercCult`% cultures containing evidence that leaders were of high social status. To explore factors that distinguished these records from those with no evidence for status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see SI for details). We then fit an elasticnet logistic regression model (with lasso penalty) of high status (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provide evidence of high status as a quality of leaders. These included *`r rev(highstatus_plot$data$Word)[1]`* and *`r rev(highstatus_plot$data$Word)[2]`*, which were the terms with the strongest predictive values; *`r rev(highstatus_plot$data$Word)[3]`* and *`r rev(highstatus_plot$data$Word)[4]`* were also strong predictors and *`r rev(highstatus_plot$data$Word)[5]`* and *`r rev(highstatus_plot$data$Word)[6]`* were also positive predictors. Notable terms which offered predictive value but with near zero coefficients included, *influence*, *authority*, *headman*, and *chief*. See Figure \@ref(fig:highstatusplot).

We then investigated if the proportion of text records with evidence for high status differed among the different levels of social and political organization, subsistence strategies, continental regions, and leader sex (again using a logistic mixed effects model with these variables as fixed effects and authors nested within cultures as random intercepts). There were modest differences. Hunter-gatherers and pastoralists societies were more likely to provide evidence for high status, as was South America and East Asia. Female leaders were also more likely to be identified as high status. Somewhat counter-intuitively, the proportion of text records indicating high status was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus). For the table of coefficients, see Supplementary Table X.

@@

### Disentangling the relationship between leadership and high status

Relationships between leadership and high status, including dependencies and causality are puzzling and currently the focus of much work in evolutionary leadership studies [e.g., @van_vugt_dual_2019; @smith_leadership_2020; @cheng_why_2020; @cheng_dominance_2019; @von_rueden_dynamics_2019]. Drawing on a comparative-evolutionary perspective, incorporating findings from nonhuman social species, @van_vugt_dual_2019 emphasize that although leadership and high social status do indeed often covary, leaders are not necessarily high-status individuals, exemplified when low-ranking individuals in nonhuman species emerge as leaders of group movement or when younger and lower status men and women initiate cooperative activities (such as raids in tribal warfare or social movements). cheng_dominance_2019 and cheng_why_2020, emphasize, however, that social hierarchies form principally in response to inter-individual variation in status, which may be gained via distinct pathways. Thus, although all leaders are not necessarily high status, especially across nonhuman species, among humans social status may be a necessary but insufficient condition for achieving and maintaing positions of leadership. In our view, social context is critical to these discussions, as is often mentioned but also under-emphasized and not well supported empirically. Processes of leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social statuses. The "low status" male who initiates a inter-group raid on a neighboring group may not be widely respected within his community or be a contending Big man, but he might be of exceptionally high status among his peers.

Our results offer some insight into relationships between leaders and social status across cultures and contexts. As mentioned, high status is the most common quality of leaders (Figure \@ref(fig:varsupport)) and the second most common benefit of leadership and a loss of status the most frequent cost (Figure \@ref(fig:varsupportcb)) mentioned in this sample of the ethnographic record which includes discussions of chiefs, work group leaders, shamans, and age-sets, from a culturally diverse sample including foraging groups and kingdoms. This lends support to the notion of high social status often being a necessary condition of leader emergence [@cheng_why_2020]. In predicting the ethnographic texts which described leaders as high status, leaders of kin groups and leaders among hunter-gatherers and pastoralists were identified as more likely to be high status relative to some other group and subsistence types. Although discussions of female leaders were rare, the presence of female leaders were also associated with high social status (see Figure \@ref(fig:groupstatus)). Interestingly, political leaders were less likely to be described as high status, which could be related to institutionalized and overt or ascribed social distinctions. Nonetheless, these findings question the informed speculation by @smith_leadership_2020 that, "leadership may therefore be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." However, our text analytic results, which relied on the semantic context of text records to predict the presence of high status leaders generally identified terms related to status and prestige (i.e., *respect*, *prestige*, *honor*) and possibly dominance or proscribed hierarchy (i.e., *rank*, *position*), as stronger predictors of high status leader than terms associated with leadership per se (i.e., *influence*, *authority*, *headman*), which lends some support to the distinction between status and leadership highlighted by @smith_leadership_2020 (though we caution from drawing strong interpretations from these text-analytic results, see Figure \@ref(fig:highstatusplot)). Overall, these results suggest associations between leadership and high status are ubiquitous across human societies, including among kin groups and among hunter-gatherers and that while effective leaders across a wide range of social contexts are the recipients of increased social status for their valuable services, ineffective leadership is associated with reduced social status.

@@

Leadership and high social status are expected to be associated with particular suites of characteristics, which potentially vary by cultural or social context, personality, or via other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @van_vugt_dual_2019]. Exploratory cluster analyses reported here suggest multiple leadership phenotypes.

@@

## Disentangling the relationship between leadership and high status, continued
